The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-06 kalman_walk_paper.tex
added citations for empirical allele-specific expression data for flycatchers and baboons

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatari2015simple}) looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless, these measures likely give us an idea of the scale of variability.
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citet{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $A$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.
\plr{JOSH: insert a citation or two to something non-human?}

@@

Is there evidence that this is actually occurring?
System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring},
and there is often significant regulatory variation segregating within populations.
Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}, and hybrid incompatibilities have been attributed to cryptic molecular divergence underlying conserved body plans \citep{gavin2013embryonic}. 
Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
%strongly reproductively incompatible species
genetically distinct species
may be nearly morphologically indistinguishable.

>>>>

It is not a new observations that there is often more than one way to do the same thing,
and that this may lead to speciation.
For instance, \citet{tulchinsky2014hybrid} (and similarly \citet{khatri2015simple}) looked for hybrid incompatibility
following regulatory sequence drift,
and \citet{porter2002speciation} studied speciation in populations expressing a simple regulatory cascade.
Both found rapid speciation following directional selection,
but only equivocal support for speciation under models of purely neutral drift.
Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
Our model differs from previous work in that we use linear systems theory
to both explore a much richer class of regulatory network models,
and provide analytical expectations in large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs to be occupied for transcription to begin),
and variation in expression level measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless, these measures likely give us an idea of the scale of variability.
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citet{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
Studying natural flycatcher populations, \citet{wang2017bayesian} found that 7.2--8.5\% of transcripts showed allele-specific expression -- indicitive of standing genetic \emph{cis}-regulatory variation. Similarly, in wild baboons, \citet{tung2015genetic} observed allele-specific expression in 23.4\% of genes studied. 
Taken together, this suggests that variation in the entries of $A$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.

@@

Is there evidence that this is actually occurring?
System drift and network rewiring has been inferred across the tree of life \citep{wotton2015quantitative, crombach2016gap, dalal2017transcription, johnson2017rewiring, ali2017quantitative},
and there is often significant regulatory variation segregating within populations.
Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association, mack2016gene}, and hybrid incompatibilities have been attributed to cryptic molecular divergence underlying conserved body plans \citep{gavin2013embryonic}. 
Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
%strongly reproductively incompatible species
genetically distinct species
may be nearly morphologically indistinguishable.
